### 2024-02-23 ms.qmd
Updates

Behavioural flexibility describes the ability to adjust behaviour adaptively in response to internal or external changes by problem-solving, innovation, or alterations in one behaviour in favour of another [@brown2010behavioral], is considered crucial for invasion success [@wright2010behavioral; @chapple2012can; @szabo2020can]. Generally, species with higher behavioural flexibility are more likely to survive and establish in new habitats because behaving flexibly enhances their capacity to exploit new resources and face new challenges [@wright2010behavioral; @chapple2012can; @szabo2020can]. For instance, when closely related species of birds were compared, successful invaders showed a higher frequency of foraging innovations [@sol2000behavioural]. According to the adaptive flexibility hypothesis, behavioural flexibility is most relevant during introduction, establishment, and spreading stages of the population into new habitats [@wright2010behavioral; @chapple2012can; @szabo2020can]. These stages can involve more than one generation, so selection may act on behavioural flexibility and the underlying mechanisms in successful invaders to ensure the persistence of behavioural flexibility through generations. In mammals, birds, reptiles and amphibians, relative brain size is a good predictor of behavioural flexibility and invasive success [@sol2000behavioural; @sol2002behavioural; @sol2008brain; @amiel_smart_2011]. Since the brain is especially sensitive to environmental changes during the early stages of development [@zhu_prenatal_2004] it is possible that the new prenatal environment faced by establishing populations can affect the behavioural flexibility of future generations [see @jonson1976effect; @lui2017chronic], with a consequent impact on invasion success. As such, we find that studying the effect of developmental conditions on behavioural flexibility can be key to understanding and predicting invasive species' success.  

@@

#### Experimental design
For this study, we used juveniles of _L. delicata_ and _L. guichenoti_ from a breeding colony kept in the lab. Established in 2019, this colony originates from animals collected in Sydney, likely from similar populations that have become invasive on Lord Howe Island [@chapple2016origin] (details of breeding colony and husbandry in Supplementary Material). To test the effect of prenatal CORT and temperature, eggs were topically supplied with 5µL of CORT dissolved in 100% Ethanol (CORT treatment) or the same volume of the vehicle (Control), and were then incubated at different temperatures (Cold -- 23ºC ± 3ºC or Hot -- 30ºC ± 3ºC) until hatching.  
To estimate behavioural flexibility, we examined lizards' performance in a reversal learning task using colour as the primary conditioned stimulus and food (frozen small *A. domestica*) as a reward (see @fig-Methods). The learning test was divided into three main phases: (1) _habituation_ -  lizards were conditioned to eat from white 3D-printed PLA ramps (9 L x 4 W x 5 H cm); (2) _association_ - lizards were conditioned to associate food with one colour (Fig. 1 Associative task); and (3) _reversal_ - the colour associated with the food was changed (Fig. 1 Reversal task). For the associative and reversal phases we used three different ramps that were identical in size and shape to the ones for the habituation phase. The colours of the ramps were green, red, and blue, as previous studies demonstrate that squamates can discriminate between these colours [@Baden_Osorio_2019_Vert_vision]. To control for potential colour biases, we assigned one group of lizards to associate the blue ramp with the accessible food in the associative task and the red ramp in the reversal, while the other group was assigned red as correct for the associative and blue for the reversal. In all trials, the position of the feeders was changed randomly to ensure subjects were using colour rather than spatial cues for the association. To avoid the influence of chemical cues, we placed the food reward in the wells of all three ramps, but in two of the wells the cricket was covered by immovable small white lids with holes on the top. Lizards were tested once daily between 11 to 12 am, when lizards are active. The habituation phase lasted 38 days, the association phase 35, and the reversal phase 40 days. We recorded and analysed manually all trials form the association and reversal phases using a standard video player (IINA). Videos were analysed by one person (P.R.) who was blind to the treatment groups. We used as the main variable whether the animal chose the correct ramp first or not. We considered that a choice was made if the head of the lizard was inside the well of one of the ramps. Lizards were considered to make a correct choice if they chose the ramp with the accessible food first. We scored lizards that chose ramps with inaccessible foods first to have made an incorrect choice. We considered a trial failed if there was no choice in one hour of recording and those trials were considered as ‘non data’ in the analyses. Here, we analysed the data of the reversal as an index of behavioural flexibility [@brown2010behavioral].

@@

#### Statistical analyses

We performed the analyses for each species separately. We fitted a total of two different Bayesian multilevel models using the brm function from the brms package (@burkner2017brms) in Quarto (@Allaire_Quarto_2022). We had per each model four parallel chains of 3000 iterations, with a warm-up interval of 1000 iterations. We used Choice - whether the individual chose correct (1) or not (0) - as the response variable, and trial, hormone (CORT versus Control), incubation temperature (Cold versus Hot), and their interaction as fixed factors. Lizard identity was employed as a random intercept, and as a random slope, we included the trial within each level of lizard identity. We used the posterior distributions of these models to evaluate differences in learning rate between treatments and species. Learning slopes were obtained using the estimates of the variable trial per each level of the hormone-temperature interaction (Treatments); values greater than zero were considered as evidence of learning, while those less or equal to zero were not. We used the pmcmc method to test whether those slopes or the comparisons between treatments (e.g. slope for CORT-Cold lizards minus CORT-Hot lizards) or species differed from zero (two-tailed tests). We considered statistical significance if p-value < 0.05.

@@

pvalue <- format_dec(sapply(estimates, pmcmc), 4)

@@

**Change RESULTS????** We found that the learning slopes for all treatments and species were higher from zero, but we did not find any difference between treatments or species. There were not significant differences between treatments in _L. delicata_ (Control-Cold - CORT-Cold = `r format_dec(mean(deli_ControlCold - deli_CORTCold), 3)`, p- value = `r format_dec(pmcmc(deli_ControlCold - deli_CORTCold), 3)`; Control-Hot - CORT-Hot = `r format_dec(mean(deli_ControlHot - deli_CORTHot), 3)`, p- value = `r format_dec(pmcmc(deli_ControlHot - deli_CORTHot), 3)`; Control-Hot - Control-Cold = `r format_dec(mean(deli_ControlHot - deli_ControlCold), 3)`, p-value = `r format_dec(pmcmc(deli_ControlHot - deli_ControlCold), 3)`; CORT-Hot - CORT-Cold = `r format_dec(mean(deli_CORTHot - deli_CORTCold), 3)`, p-value = `r format_dec(pmcmc(deli_CORTHot - deli_CORTCold), 3)`) (see @fig-results) or _L. guichenoti_ (Control-Cold - CORT-Cold = `r format_dec(mean(guich_ControlCold - guich_CORTCold), 3)`, p- value = `r format_dec(pmcmc(guich_ControlCold - guich_CORTCold), 3)`; Control-Hot - CORT-Hot = `r format_dec(mean(guich_ControlHot - guich_CORTHot), 3)`, p- value = `r format_dec(pmcmc(guich_ControlHot - guich_CORTHot), 3)`; Control-Hot - Control-Cold = `r format_dec(mean(guich_ControlHot - guich_ControlCold), 3)`, p-value = `r format_dec(pmcmc(guich_ControlHot - guich_ControlCold), 3)`; CORT-Hot - CORT-Cold = `r format_dec(mean(guich_CORTHot - guich_CORTCold), 3)`, p-value = `r format_dec(pmcmc(guich_CORTHot - guich_CORTCold), 3)`) (see @fig-results). When pooled together, we did not find any significant effect of incubation temperature (_L. delicata_: Hot learning slope - Cold learning slope = `r format_dec(mean((deli_CORTHot + deli_ControlHot) - (deli_CORTCold + deli_ControlCold)),3)`, p-value = `r format_dec(pmcmc((deli_CORTHot + deli_ControlHot) - (deli_CORTCold + deli_ControlCold)),3)`; _L. guichenoti_: Hot learning slope - Cold learning slope = `r format_dec(mean((guich_CORTHot + guich_ControlHot) - (guich_CORTCold + guich_ControlCold)),3)`, p-value = `r format_dec(pmcmc((guich_CORTHot + guich_ControlHot) -(guich_CORTCold + guich_ControlCold)),3)`) or the hormone (_L. delicata_: Control learning slope - CORT learning slope = `r format_dec(mean((deli_ControlCold + deli_ControlHot) - (deli_CORTCold + deli_CORTHot)), 3)`, p-value = `r format_dec(pmcmc((deli_ControlCold + deli_ControlHot) - (deli_CORTCold + deli_CORTHot)), 3)`; _L. guichenoti_: Control learning slope - CORT learning slope = `r format_dec(mean((guich_ControlCold + guich_ControlHot)-(guich_CORTCold + guich_CORTHot)), 3)`, p-value = `r format_dec(pmcmc((guich_ControlCold + guich_ControlHot)-(guich_CORTCold + guich_CORTHot)), 3)`). Finally, there were not significant differences when we compared the estimated slopes between species (_L. delicata_ - _L. guichenoti_ = `r format_dec(mean((deli_CORTCold + deli_ControlCold + deli_CORTHot + deli_ControlHot) - (guich_CORTCold + guich_ControlCold + guich_CORTHot + guich_ControlHot)), 3)`, p-value = `r format_dec(pmcmc((deli_CORTCold + deli_ControlCold + deli_CORTHot + deli_ControlHot) - (guich_CORTCold + guich_ControlCold + guich_CORTHot + guich_ControlHot)), 3)`).

@@

pvalue_2 <- format_dec(sapply(estimates_2, pmcmc), 4)

>>>>

Behavioural flexibility describes the ability of individuals to adjust behaviour adaptively in response to internal or external changes in the environment [@brown2010behavioral]. It comprises a suite of behavioural aspects like problem-solving, innovation, or alterations in one behaviour in favour of another [@brown2010behavioral]; and it is predicted to be crucial for invasion success [@wright2010behavioral; @chapple2012can; @szabo2020can]. Generally, species with greater behavioural flexibility are more likely to survive and establish in new habitats because behaving flexibly enhances their capacity to exploit new resources and face new challenges [@wright2010behavioral; @chapple2012can; @szabo2020can]. For instance, in closely related bird species successful invaders show a higher frequency of foraging innovations [@sol2000behavioural]. According to the 'adaptive flexibility hypothesis', behavioural flexibility is most relevant during introduction, establishment, and spread of an invasive population [@wright2010behavioral; @chapple2012can; @szabo2020can]. In mammals, birds, reptiles and amphibians, relative brain size has been shown to be a good predictor of behavioural flexibility and invasive success [@sol2000behavioural; @sol2002behavioural; @sol2008brain; @amiel_smart_2011]. However, the brain is also sensitive to environmental changes during the early stages of development [@zhu_prenatal_2004] creating uncertainty about how environmental insults faced by early invaders might impact behavioural flexibility and thus the probability of invasion success.    

@@

#### Manipulating Early Thermal and CORT Environments  
To manipulate early thermal environment and CORT, we collected eggs of _L. delicata_ and _L. guichenoti_ from a lab breeding colony that was established from a Sydney population in 2019. The Sydney population is thought to be the source population for invasive populations established on Lord Howe Island [@chapple2016origin]. Eggs were topically supplied with 5µL of CORT dissolved in 100% Ethanol (CORT treatment) or the same volume of the vehicle (Control) and were then incubated at different temperatures (Cold – 23ºC ± 3ºC or Hot – 30ºC ± 3ºC) until hatching in a fully factorial design. For full details of breeding colony and husbandry see Supplementary Material.

#### Quantifying Behavioural Flexibility  
Briefly, the learning test was divided in three main phases: a habituation phase where lizards had to learn to eat a frozen-thawed cricket from white 3D-printed PLA ramps (9 L x 4 W x 5 H cm). The habituation phase lasted 38 days. This was followed by a colour association task where lizards had to associate the food reward with one of two colours (red or blue) (@fig-Methods Associative task). The associative learning task lasted 35 days. Finally, to quantify behavioural flexibility, we reversed the previously trained colour (red or blue) so that they needed to learn to associate the food with the opposite colour (@fig-Methods Reversal task). The reversal learning task lasted a total of 40 days. Lizards were tested once daily between 11 to 12 am when they were most active.  
In associative and reversal tasks we used three different ramps that were identical in size and shape to the ones for the habituation phase. The colours of the ramps were green, red, and blue, as previous studies demonstrate that squamates can discriminate between these colours [@Baden_Osorio_2019_Vert_vision]. To avoid the influence of chemical cues, we placed the food reward in all three ramps, but in two of them the cricket was covered by immovable small lids with holes on the top. To control for potential colour biases, we assigned one group of lizards within each treatment to associate food in the blue ramp followed by the red ramp in the reversal, while in the other group the order was reversed. In all trials, the position of the feeders was changed randomly to ensure subjects were using colour rather than spatial cues for the association.  
We recorded whether the animal chose the correct ramp first or not. We considered that a choice was made if the head of the lizard was inside the well of one of the ramps. We considered a trial failed if there was no choice in one hour of recording and those trials. Here, we analysed the data of the reversal task as a measure of behavioural flexibility [@brown2010behavioral].

@@

#### Statistical analyses  
We performed the analyses for each species separately. We fitted a total of two different Bayesian multilevel models using the brm function from the _brms_ package [@burkner2017brms] using the R extension (version 2.8.2) [@R] for Visual Studio Code. We run four parallel MCMC chains of 3000 iterations for each model, with a warmup period of 1000 iterations. We used Choice - whether the individual chose correct (1) or not (0) - as the response variable, and trial, hormone (CORT versus Control), incubation temperature (Cold versus Hot), and their interaction as fixed factors. We included a random intercept and slope (trial) for each lizard in our models. We used the posterior distributions of parameters from these models to test for differences in learning rate between treatments and species. Learning slopes were obtained using the estimates of the variable trial per each level of the hormone-temperature interaction (Treatments); values greater than zero were considered as evidence of learning, while those less or equal to zero were not. We used the pmcmc method to test whether those slopes or the comparisons between treatments (e.g. slope for CORT-Cold lizards minus CORT-Hot lizards) or species differed from zero (two-tailed tests). We considered statistical significance if p-value < 0.05.

@@

pvalue <- format_p(sapply(estimates, pmcmc), 4)

@@

 We found that the learning slopes for all treatments and species were higher from zero (@tbl-data), but we did not find any difference between treatments or species in the learning rate (@fig-results). There were not significant differences between treatments in _L. delicata_ (Control-Cold - CORT-Cold = `r format_dec(mean(deli_ControlCold - deli_CORTCold), 3)`, p- value = `r format_dec(pmcmc(deli_ControlCold - deli_CORTCold), 3)`; Control-Hot - CORT-Hot = `r format_dec(mean(deli_ControlHot - deli_CORTHot), 3)`, p- value = `r format_dec(pmcmc(deli_ControlHot - deli_CORTHot), 3)`; Control-Hot - Control-Cold = `r format_dec(mean(deli_ControlHot - deli_ControlCold), 3)`, p-value = `r format_p(pmcmc(deli_ControlHot - deli_ControlCold), 3)`; CORT-Hot - CORT-Cold = `r format_dec(mean(deli_CORTHot - deli_CORTCold), 3)`, p-value = `r format_p(pmcmc(deli_CORTHot - deli_CORTCold), 3)`) (see @fig-results) or _L. guichenoti_ (Control-Cold - CORT-Cold = `r format_dec(mean(guich_ControlCold - guich_CORTCold), 3)`, p- value = `r format_dec(pmcmc(guich_ControlCold - guich_CORTCold), 3)`; Control-Hot - CORT-Hot = `r format_dec(mean(guich_ControlHot - guich_CORTHot), 3)`, p- value = `r format_dec(pmcmc(guich_ControlHot - guich_CORTHot), 3)`; Control-Hot - Control-Cold = `r format_dec(mean(guich_ControlHot - guich_ControlCold), 3)`, p-value = `r format_p(pmcmc(guich_ControlHot - guich_ControlCold), 3)`; CORT-Hot - CORT-Cold = `r format_dec(mean(guich_CORTHot - guich_CORTCold), 3)`, p-value = `r format_p(pmcmc(guich_CORTHot - guich_CORTCold), 3)`) (see @fig-results).  
 We compared each factor (hormone and incubation temperature) separately by pooling together the observations for all the individuals from the same hormone treatment or the same incubation temperature. In this regard, we did not find any significant effect of the hormone treatment (_L. delicata_: Control learning slope - CORT learning slope = `r format_dec(mean((deli_ControlCold + deli_ControlHot) - (deli_CORTCold + deli_CORTHot)), 3)`, p-value = `r format_p(pmcmc((deli_ControlCold + deli_ControlHot) - (deli_CORTCold + deli_CORTHot)), 3)`; _L. guichenoti_: Control learning slope - CORT learning slope = `r format_dec(mean((guich_ControlCold + guich_ControlHot)-(guich_CORTCold + guich_CORTHot)), 3)`, p-value = `r format_p(pmcmc((guich_ControlCold + guich_ControlHot)-(guich_CORTCold + guich_CORTHot)), 3)`) or the incubation temperature (_L. delicata_: Hot learning slope - Cold learning slope = `r format_dec(mean((deli_CORTHot + deli_ControlHot) - (deli_CORTCold + deli_ControlCold)),3)`, p-value = `r format_p(pmcmc((deli_CORTHot + deli_ControlHot) - (deli_CORTCold + deli_ControlCold)),3)`; _L. guichenoti_: Hot learning slope - Cold learning slope = `r format_dec(mean((guich_CORTHot + guich_ControlHot) - (guich_CORTCold + guich_ControlCold)),3)`, p-value = `r format_p(pmcmc((guich_CORTHot + guich_ControlHot) -(guich_CORTCold + guich_ControlCold)),3)`) in either species.  
 Finally, when all individuals from all the treatments were pooled together and species were compared, there were not significant differences in the estimated slopes (_L. delicata_ - _L. guichenoti_ = `r format_dec(mean((deli_CORTCold + deli_ControlCold + deli_CORTHot + deli_ControlHot) - (guich_CORTCold + guich_ControlCold + guich_CORTHot + guich_ControlHot)), 3)`, p-value = `r format_p(pmcmc((deli_CORTCold + deli_ControlCold + deli_CORTHot + deli_ControlHot) - (guich_CORTCold + guich_ControlCold + guich_CORTHot + guich_ControlHot)), 3)`).

@@

pvalue_2 <- format_p(sapply(estimates_2, pmcmc), 4)
